---
title: "Multicollinearity and Misspecification"
author: "Grok 3 :P"
date: 'Sys.Date()'
---

# Computational Experiment: Exploring Violations of Gauss-Markov Assumptions

This notebook lets you explore four key violations of the Gauss-Markov assumptions using synthetic data:  
1. Multicollinearity  
2. Under-specification (Omitted Variable Bias)  
3. Over-specification  
4. Functional Form Misspecification  

You’ll generate data, run regressions, and use tests and plots to detect these issues. Interpret the results using the hypotheses below—no hand-holding here!

## Hypotheses for Each Test

- **Multicollinearity**:  
  - *VIF*: Variance Inflation Factor. VIF > 10 suggests multicollinearity inflates coefficient variance.  
  - *Correlation Matrix*: Correlations > 0.8 or < -0.8 between regressors indicate multicollinearity.  
  - *Condition Number*: > 30 suggests multicollinearity in the model matrix.  

- **Under-specification (Omitted Variable Bias)**:  
  - No direct test. High residuals or a big fit improvement when adding omitted variables suggest an issue.  

- **Over-specification**:  
  - Adding irrelevant variables doesn’t bias coefficients but inflates standard errors. Compare standard errors with the true model.  

- **Functional Form Misspecification**:  
  - *Ramsey RESET Test*: Significant p-value (< 0.05) suggests missing nonlinear terms.  
  - *Residual Plots*: Patterns (e.g., curves) indicate the functional form is wrong.
  
```{r}
# Load required packages
library(lmtest)  # For Ramsey RESET test
library(car)     # For VIF
```

*Data Generation*: `generate_data()` creates synthetic data with toggles for each violation. It’s flexible and keeps the noise manageable.

```{r}
# Generate synthetic data with optional violations
generate_data <- function(n = 100, beta = c(1, 2, 3), multicol = FALSE, under_spec = FALSE, 
                         over_spec = FALSE, nonlinear = FALSE) {
  # Generate regressors
  if (multicol) {
    x1 <- rnorm(n)
    x2 <- x1 + rnorm(n, sd = 0.1)  # High correlation with x1
  } else {
    x1 <- rnorm(n)
    x2 <- rnorm(n)
  }
  
  if (over_spec) {
    x3 <- rnorm(n)  # Irrelevant variable
  }
  
  # Generate response variable
  if (nonlinear) {
    y <- beta[1] + beta[2] * x1 + beta[3] * x1^2 + rnorm(n)
  } else {
    y <- beta[1] + beta[2] * x1 + beta[3] * x2 + rnorm(n)
  }
  
  # Create data frame
  data <- data.frame(y = y, x1 = x1, x2 = x2)
  if (over_spec) data$x3 <- x3
  if (under_spec) data <- data[, !names(data) %in% "x2"]
  
  return(data)
}
```

*Experiment*: `run_experiment()` runs the regression, performs relevant tests, and generates plots. It’s modular—each violation gets its own checks.

```{r}
# Run regression and tests for a given scenario
run_experiment <- function(data, multicol = FALSE, under_spec = FALSE, over_spec = FALSE, 
                          nonlinear = FALSE, true_model = FALSE) {
  # Define the regression model based on scenario
  if (true_model) {
    model <- lm(y ~ x1 + x2, data = data)
  } else if (under_spec) {
    model <- lm(y ~ x1, data = data)
  } else if (over_spec) {
    model <- lm(y ~ x1 + x2 + x3, data = data)
  } else {
    model <- lm(y ~ x1 + x2, data = data)
  }
  
  # Print model summary
  cat("\nModel Summary:\n")
  print(summary(model))
  
  # Multicollinearity tests
  if (multicol) {
    cat("\nMulticollinearity Tests:\n")
    cat("VIF:\n")
    print(vif(model))
    cat("Correlation Matrix:\n")
    print(cor(data[, !names(data) %in% "y"]))
    cat("Condition Number:", kappa(model.matrix(model)), "\n")
    plot(data$x1, data$x2, main = "x1 vs x2", xlab = "x1", ylab = "x2")
  }
  
  # Functional form misspecification tests
  if (nonlinear) {
    cat("\nFunctional Form Misspecification Tests:\n")
    cat("Ramsey RESET Test:\n")
    print(resettest(model, power = 2:3))
    plot(fitted(model), residuals(model), main = "Residuals vs Fitted", 
         xlab = "Fitted Values", ylab = "Residuals")
    abline(h = 0, col = "red")
  }
  
  # For under-specification: residual plot
  if (under_spec) {
    plot(fitted(model), residuals(model), main = "Residuals vs Fitted (Under-specified)", 
         xlab = "Fitted Values", ylab = "Residuals")
    abline(h = 0, col = "red")
  }
}
```

```{r}
# Scenario 1: Multicollinearity
cat("\n=== Multicollinearity Scenario ===\n")
data_multicol <- generate_data(multicol = TRUE)
run_experiment(data_multicol, multicol = TRUE)

# Scenario 2: Under-specification
cat("\n=== Under-specification Scenario ===\n")
data_under <- generate_data(under_spec = TRUE)
run_experiment(data_under, under_spec = TRUE)

# Scenario 3: Over-specification
cat("\n=== Over-specification Scenario ===\n")
data_over <- generate_data(over_spec = TRUE)
run_experiment(data_over, over_spec = TRUE)

# Scenario 4: Functional Form Misspecification
cat("\n=== Nonlinear Scenario ===\n")
data_nonlinear <- generate_data(nonlinear = TRUE)
run_experiment(data_nonlinear, nonlinear = TRUE)

# Scenario 5: Correct Model (Baseline)
cat("\n=== Correct Model Scenario ===\n")
data_correct <- generate_data()
run_experiment(data_correct, true_model = TRUE)
```


## Test it yourself

- *Multicollinearity*: Check if VIFs are huge or if $x_1$ and $x_2$ look like a straight line in the scatter plot.

- *Under-specification*: Look for patterns in the residual plot—random scatter is good, curves are bad.

- *Over-specification*: Compare standard errors with the correct model. Bigger errors? You’ve got junk in there.

- *Nonlinear*: RESET p-value < 0.05 or a curvy residual plot screams “wrong form!”
